---
title: "Leaflet + Shiny"
subtitle: "PHS 7045: Advanced Programming"
author: "Haojia Li, MS"
format:
    revealjs:
      embed-resources: true
      slide-number: true
      scrollable: true
      smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F)
slides_eval <- TRUE
```

# Intro

## Kick off

::: {style="font-size: 20pt"}

To wrap our heads around things, let's start by diving into the [COVID-19 tracker app](https://shiny.rstudio.com/gallery/covid19-tracker.html) and getting a feel for what it's all about!

:::

## Plan for today

::: {style="font-size: 20pt"}

What we'll be using

1. **Leaflet**: "Create Interactive Web Maps with the JavaScript 'Leaflet' Library" ([read more](https://cran.r-project.org/web/packages/leaflet/)).

2. **plotly**: "Create Interactive Web Graphics via 'plotly.js'" ([read more](https://cran.r-project.org/web/packages/plotly/))

3. **DT**: "A Wrapper of the JavaScript Library 'DataTables'" ([read more](https://cran.r-project.org/web/packages/DT/))

4. **shiny**: "Web Application Framework for R" ([read more](https://cran.r-project.org/web/packages/shiny/))

We will adopt the the practice in Part 3, [Lab 3](https://uofuepibio.github.io/PHS7045-advanced-programming/week-03-lab.html), visualizing the representative USAF station(s) for weather conditions.

:::

# Data preparation

## 1. Read-in and data combination

::: {style="font-size: 20pt"}

The met data contains the records of weather conditions, including wind speed, temperature, dew point, atmospheric pressure, and relative humidity, at USAF stations in August 2019.

```{r}
# ---- load data.table package ----
library(data.table)

# ---- set up weathers ----
weathers_var <- c("wind.sp", "temp", "dew.point", "atm.press", "rh")
weathers_label <- c("Wind speed", "Temperature", "Dew point", "Atmospheric pressure", "Relative humidity")
weathers_unit <- c(" km/h", "&#8451;", "&#8451;", " hPa", "%")

# ---- met data read-in and processing ----
met <- fread("https://raw.githubusercontent.com/USCbiostats/data-science-data/master/02_met/met_all.gz")
# select columns
met <- met[, c("USAFID", "day", "hour", weathers_var), with = F]

# ---- station data read-in and processing ----
stations <- fread("ftp://ftp.ncdc.noaa.gov/pub/data/noaa/isd-history.csv")
# filter stations in US and select columns
stations <- stations[CTRY == "US", .(USAF, `STATION NAME`, STATE, LON, LAT)]

# drop NAs (999999) in USAF
stations[, USAF := as.integer(USAF)]
stations[, USAF := fifelse(USAF == 999999, NA_integer_, USAF)]
stations <- stations[!is.na(USAF)]

# drop NAs in STATE
stations[, STATE := fifelse(STATE == "", NA_character_, STATE)]
stations <- stations[!is.na(STATE)]

# keep only one data point for each USAF
stations[, n := 1:.N, by = .(USAF)]
stations <- stations[n == 1,][, n := NULL]

# ---- merge station into met ----
met <- merge(met, stations, by.x = "USAFID", by.y = "USAF", all.x = T)
```

:::

## 2. Data aggregation

::: {style="font-size: 20pt"}

We want to calculate the median value of weather conditions at different levels: station, state, and country.

```{r}
# number of records and median value of weathers at station level
met_station <- met[, c(.N, lapply(.SD, median, na.rm = T)), 
                   by = .(USAFID, STATE, `STATION NAME`, LON, LAT),
                   .SDcols = weathers_var]

# median value of weathers at state level
met_state <- met[, lapply(.SD, median, na.rm = T), by = STATE,.SDcols = weathers_var]
# merge state-level median values
met_station <- merge(met_station, met_state, by = "STATE", suffixes = c(".station", ".state"))

# median value of weathers at country level
met_us <- met[, lapply(.SD, median, na.rm = T), .SDcols = weathers_var]
# add country-level median values
colnames(met_us) <- paste0(colnames(met_us), ".us")
met_station <- cbind(met_station, met_us)

library(kableExtra)
kable(met_station, row.names = T) |> kable_styling(font_size = 12) |> scroll_box(width = "100%", height = "150px")
```

:::

## 3. Identify representative station(s) 

::: {style="font-size: 20pt"}

Write a function to identify the representative station(s), at state or country level,  by euclidean distance. User can specify the weather condition(s) of interest and number and/or percentage of the representative.

```{r}
sta_represent <- function(
    weathers = weathers_var, # weather conditions of interest
    top_n = 3, # number of representative stations
    top_perc = 10, # upper limit of percentage for representatives
    level = c("state", "us") # at which level to identify representatives
) {
  weathers <- match.arg(weathers, several.ok = T)
  level <- match.arg(level)
  
  fdata <- copy(met_station)
  
  # calculate euclidean distance
  fdata$weather_dist <- sqrt(rowSums(
    (as.matrix(fdata[, paste0(weathers, ".", "station"), with = F]) - 
       as.matrix(fdata[, paste0(weathers, ".", level), with = F])) ^ 2,
  ))
  
  # identify the representative station by distance and number of records
  if(level == "state")
    res <- fdata[order(weather_dist, -N), head(.SD, min(top_n, ceiling(.N * top_perc/100))), 
                 by = STATE]
  if(level == "us")
    res <- fdata[order(weather_dist, -N), head(.SD, min(top_n, ceiling(.N * top_perc/100)))]
  
  # add text to be used as popup
  res$text <- paste0(
    "USAFID: ", res$USAFID, "<br>", 
    "Name: ", res$`STATION NAME`, "<br>",
    "Number of records: ", res$N, "<br>",
    sapply(weathers, \(x) {
      paste0(weathers_label[match(x, weathers_var)], ": ",
             unlist(res[, paste0(x, ".station"), with = F]) |> round(1),
             weathers_unit[match(x, weathers_var)], "<br>")
    }) |> apply(1, paste, collapse = ""),
    "Distance score: ", round(res$weather_dist, 1)
  )
  
  # drop missing values
  res <- res[!is.na(weather_dist), ]
  return(res)
}
```

:::

# Leaflet: Create Interactive Web Maps

## 0. Overview

::: {style="font-size: 20pt"}

[Leaflet](https://leafletjs.com/) is one of the most popular open-source JavaScript libraries for interactive maps.

The [`leaflet`](https://rstudio.github.io/leaflet/) R package makes it easy to integrate and control Leaflet maps in R.

You can use this package at the R console, within R Markdown documents, and within Shiny applications.

```{r}
# load leaflet
if(!require("leaflet", quietly = TRUE)) install.packages("leaflet")
library(leaflet)
```

There are several forms of spatial data which can be processed by leaflet. We will only focus on data frame with lng/lat columns from base R.

The essential steps to build a Leaflet map include: 

1. Create a map widget by calling `leaflet()`;

2. Set up basemaps by `addTiles()` or `addProviderTiles()`;

3. Manipulate the attributes of the map widget using methods such as `setView()`;

3. Add components, such as `addMarkers()`, `addCircles()`, and `addLayersControl()`;

4. Print the map widget to display it.

:::

## 1. Create a map widget

::: {style="font-size: 20pt"}

We can create a map widget by simply calling `leaflet()`.

```{r out.width="100%"}
m <- leaflet()
m
```

:::

## 2. Set up basemaps by `addTiles()`

::: {style="font-size: 20pt"}

The easiest way to add tiles is by calling `addTiles()` with no arguments; by default, [OpenStreetMap](https://www.openstreetmap.org/#map=2/13.2/-113.3) tiles are used.

```{r, out.width="100%", out.height="400px"}
m |> addTiles()
```

:::

## 2. Set up basemaps by `addProviderTiles()`

::: {style="font-size: 20pt"}

Alternatively, many popular free third-party basemaps can be added using the `addProviderTiles()` function, which is implemented using the leaflet-providers plugin. See [here](http://leaflet-extras.github.io/leaflet-providers/preview/index.html) for the complete set. You can use `names(providers)` to view all of the options.

```{r, out.width="100%", out.height="400px"}
m <- leaflet() |> addProviderTiles(providers$CartoDB.Positron)
m
```

:::

## 3. Map methods

::: {style="font-size: 20pt"}

- `setView()` sets the center of the map view and the zoom level;
- `fitBounds()` fits the view into the rectangle [lng1, lat1] – [lng2, lat2];
- `clearBounds()` clears the bound, so that the view will be automatically determined by the range of latitude/longitude data in the map layers if provided.

```{r, out.width="100%", out.height="400px"}
m |> setView(lng = -111.8272, lat = 40.7648, zoom = 17)
```

:::

# Shiny: Web Application Framework for R





